---
title: "Correcting batch effects using the mutual nearest neighbour method"
author: 
- name: Aaron T. L. Lun
  affiliation: &CRUK Cancer Research UK Cambridge Institute, Li Ka Shing Centre, Robinson Way, Cambridge CB2 0RE, United Kingdom
date: 10 December 2017
vignette: >
  %\VignetteIndexEntry{Correcting batch effects using the mutual nearest neighbour method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
output: 
    BiocStyle::html_document
bibliography: ref.bib
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
library(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)

# Setting single-core unless explicitly specified otherwise.
library(BiocParallel)
register(SerialParam())

# Deciding whether we want to re-download everything or not.
on.bioc <- FALSE

# Further arguments for local execution.
opts_chunk$set(fig.asp=1)
if (!on.bioc) {
    opts_chunk$set(dpi=300, dev="png", dev.args=list(pointsize=15))
    options(bitmapType="cairo", width=100)
}
```

```{r, eval=on.bioc, echo=FALSE, results='hide'}
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE81nnn/GSE81076/suppl/GSE81076%5FD2%5F3%5F7%5F10%5F17%2Etxt%2Egz", 
		      destfile="GSE81076_D2_3_7_10_17.txt.gz", mode="wb")
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE85nnn/GSE85241/suppl/GSE85241%5Fcellsystems%5Fdataset%5F4donors%5Fupdated%2Ecsv%2Egz", 
			  destfile="GSE85241_cellsystems_dataset_4donors_updated.csv", mode="w")
download.file("https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-5061/files/E-MTAB-5061.processed.1.zip",
			  destfile="E-MTAB-5061.processed.1.zip", mode="wb")
```

# Introduction

Large single-cell RNA sequencing (scRNA-seq) projects usually need to generate data across multiple batches due to logistical constraints.
However, the processing of different batches is often subject to uncontrollable differences, e.g., changes in operator, differences in reagent quality.
This results in systematic differences in the observed expression in cells from different batches, which we refer to as "batch effects".
Batch effects are problematic as they can be major drivers of heterogeneity in the data, masking the relevant biological differences and complicating interpretation of the results.

Computational correction of these effects is critical for eliminating batch-to-batch variation, allowing data across multiple batches to be combined for valid downstream analysis.
However, existing methods such as `removeBatchEffect()` [@ritchie2015limma] assume that the composition of cell populations are either known or the same across batches.
This workflow describes the application of an alternative strategy for batch correction based on the detection of mutual nearest neighbours [@haghverdi2018correcting].
The `mnnCorreect()` approach does not rely on pre-defined or equal population compositions across batches, only requiring that a subset of the population be shared between batches.
We demonstrate its use on three human pancreas scRNA-seq datasets from different groups and using different protocols.

# Processing the different datasets

## CEL-seq, GSE81076

### Loading in the data

This dataset was generated by @grun2016denovo using the CEL-seq protocol with unique molecular identifiers (UMIs) and ERCC spike-ins.
Count tables were obtained from the NCBI Gene Expression Omnibus using the accession number above.
We first read the table into memory.

```{r}
gse81076.df <- read.table("GSE81076_D2_3_7_10_17.txt.gz", sep='\t', 
    header=TRUE, stringsAsFactors=FALSE, row.names=1)
dim(gse81076.df)
```

Unfortunately, the data and metadata are all mixed together in this file.
As a result, we need to manually extract the metadata from the column names.

```{r}
donor.names <- sub("^(D[0-9]+).*", "\\1", colnames(gse81076.df))
table(donor.names)
plate.id <- sub("^D[0-9]+(.*)_.*", "\\1", colnames(gse81076.df))
table(plate.id)
```

Another irritating feature of this dataset is that gene symbols were supplied, rather than stable identifiers such as Ensembl.
We convert all row names to Ensembl identifiers, removing `NA` or duplicated entries (with the exception of spike-in transcripts). 

```{r}
gene.symb <- gsub("__chr.*$", "", rownames(gse81076.df))
is.spike <- grepl("^ERCC-", gene.symb)
table(is.spike)

library(org.Hs.eg.db)
gene.ids <- mapIds(org.Hs.eg.db, keys=gene.symb, keytype="SYMBOL", column="ENSEMBL")
gene.ids[is.spike] <- gene.symb[is.spike]

keep <- !is.na(gene.ids) & !duplicated(gene.ids)
gse81076.df <- gse81076.df[keep,]
rownames(gse81076.df) <- gene.ids[keep]
summary(keep)
```

We create a `SingleCellExperiment` object to store the counts and metadata together.
This reduces the risk of book-keeping errors in later steps of the analysis.
Note that we re-identify the spike-in rows, as the previous indices would have changed after the subsetting.

```{r}
library(SingleCellExperiment)
sce.gse81076 <- SingleCellExperiment(list(counts=as.matrix(gse81076.df)),
	colData=DataFrame(Donor=donor.names, Plate=plate.id),
	rowData=DataFrame(Symbol=gene.symb[keep]))
isSpike(sce.gse81076, "ERCC") <- grepl("^ERCC-", rownames(gse81076.df)) 
sce.gse81076  
```

### Quality control and normalization

We compute quality control (QC) metrics for each cell [@mccarthy2017scater] and identify cells with low library sizes, low numbers of expressed genes, or high ERCC content.

```{r}
library(scater)
sce.gse81076 <- calculateQCMetrics(sce.gse81076, compact=TRUE)
QC <- sce.gse81076$scater_qc
low.lib <- isOutlier(QC$all$log10_total_counts, type="lower", nmad=3)
low.genes <- isOutlier(QC$all$log10_total_features_by_counts, type="lower", nmad=3)
high.spike <- isOutlier(QC$feature_control_ERCC$pct_counts, type="higher", nmad=3)
data.frame(LowLib=sum(low.lib), LowNgenes=sum(low.genes), 
	HighSpike=sum(high.spike, na.rm=TRUE))
```

Cells with extreme values for these QC metrics are presumed to be of low quality and are removed.
A more thorough analysis would examine the distributions of these QC metrics beforehand, but we will skip that step for brevity here.

```{r}
discard <- low.lib | low.genes | high.spike
sce.gse81076 <- sce.gse81076[,!discard]
summary(discard)
```

We compute size factors for the endogenous genes using the deconvolution method [@lun2016pooling].
This is done with pre-clustering through `quickCluster()` to avoid pooling together very different cells.

```{r}
library(scran)
clusters <- quickCluster(sce.gse81076, min.mean=0.1)
table(clusters)
sce.gse81076 <- computeSumFactors(sce.gse81076, min.mean=0.1, clusters=clusters)
summary(sizeFactors(sce.gse81076))
```

We also compute size factors for the spike-in transcripts [@lun2017assessing].
Recall that we set `general.use=FALSE` to ensure that the spike-in size factors are only applied to the spike-in transcripts.

```{r}
sce.gse81076 <- computeSpikeFactors(sce.gse81076, general.use=FALSE)
summary(sizeFactors(sce.gse81076, "ERCC"))
```

We then compute normalized log-expression values for use in downstream analyses.

```{r}
sce.gse81076 <- normalize(sce.gse81076)
```

### Identifying highly variable genes

We identify highly variable genes (HVGs) using `trendVar()` and `decomposeVar()`.
Note that we set `block=` to ensure that uninteresting differences between plates or donors do not inflate the variance.

```{r, fig.cap="Variance of normalized log-expression values for each gene in the GSE81076 dataset, plotted against the mean log-expression. The blue line represents the mean-dependent trend fitted to the variances of the spike-in transcripts (red). The small apparent discrepancy in the fit is caused by the fact that the trend is fitted robustly to the block-wise variances of the spike-ins, while the variances shown are averaged across blocks and not robust to outliers."}
block <- paste0(sce.gse81076$Plate, "_", sce.gse81076$Donor)
fit <- trendVar(sce.gse81076, block=block, parametric=TRUE) 
dec <- decomposeVar(sce.gse81076, fit)
plot(dec$mean, dec$total, xlab="Mean log-expression", 
	ylab="Variance of log-expression", pch=16)
is.spike <- isSpike(sce.gse81076)
points(dec$mean[is.spike], dec$total[is.spike], col="red", pch=16)
curve(fit$trend(x), col="dodgerblue", add=TRUE)
```

We order genes by decreasing biological component, revealing some usual suspects such as insulin and glucagon.
We will be using this information later when performing feature selection prior to running `mnnCorrect()`.

```{r}
dec <- cbind(Symbol=rowData(sce.gse81076)$Symbol, dec)
dec[head(order(dec$bio, decreasing=TRUE)),]
```

## CEL-seq2, GSE85241

### Loading in the data

This dataset was generated by @muraro2016singlecell using the CEL-seq2 protocol with unique molecular identifiers (UMIs) and ERCC spike-ins.
Count tables were obtained from the NCBI Gene Expression Omnibus using the accession number above.
We first read the table into memory.

```{r}
gse85241.df <- read.table("GSE85241_cellsystems_dataset_4donors_updated.csv", 
    sep='\t', h=TRUE, row.names=1, stringsAsFactors=FALSE)
dim(gse85241.df)
```

We extract the metadata from the column names.

```{r}
donor.names <- sub("^(D[0-9]+).*", "\\1", colnames(gse85241.df))
table(donor.names)
plate.id <- sub("^D[0-9]+\\.([0-9]+)_.*", "\\1", colnames(gse85241.df))
table(plate.id)
```

Yet again, gene symbols were supplied instead of Ensembl or Entrez identifiers.
We convert all row names to Ensembl identifiers, removing `NA` or duplicated entries (with the exception of spike-in transcripts). 

```{r}
gene.symb <- gsub("__chr.*$", "", rownames(gse85241.df))
is.spike <- grepl("^ERCC-", gene.symb)
table(is.spike)

library(org.Hs.eg.db)
gene.ids <- mapIds(org.Hs.eg.db, keys=gene.symb, keytype="SYMBOL", column="ENSEMBL")
gene.ids[is.spike] <- gene.symb[is.spike]

keep <- !is.na(gene.ids) & !duplicated(gene.ids)
gse85241.df <- gse85241.df[keep,]
rownames(gse85241.df) <- gene.ids[keep]
summary(keep)
```

We create a `SingleCellExperiment` object to store the counts and metadata together.

```{r}
sce.gse85241 <- SingleCellExperiment(list(counts=as.matrix(gse85241.df)),
	colData=DataFrame(Donor=donor.names, Plate=plate.id),
	rowData=DataFrame(Symbol=gene.symb[keep]))
isSpike(sce.gse85241, "ERCC") <- grepl("^ERCC-", rownames(gse85241.df)) 
sce.gse85241  
```

### Quality control and normalization

We compute QC metrics for each cell and identify cells with low library sizes, low numbers of expressed genes, or high ERCC content.

```{r}
sce.gse85241 <- calculateQCMetrics(sce.gse85241, compact=TRUE)
QC <- sce.gse85241$scater_qc
low.lib <- isOutlier(QC$all$log10_total_counts, type="lower", nmad=3)
low.genes <- isOutlier(QC$all$log10_total_features_by_counts, type="lower", nmad=3)
high.spike <- isOutlier(QC$feature_control_ERCC$pct_counts, type="higher", nmad=3)
data.frame(LowLib=sum(low.lib), LowNgenes=sum(low.genes), 
	HighSpike=sum(high.spike, na.rm=TRUE))
```

Low-quality cells are defined as those with extreme values for these QC metrics and are removed.

```{r}
discard <- low.lib | low.genes | high.spike
sce.gse85241 <- sce.gse85241[,!discard]
summary(discard)
```

We compute size factors for the endogenous genes and spike-in transcripts, and use them to compute log-normalized expression values.

```{r}
clusters <- quickCluster(sce.gse85241, min.mean=0.1, method="igraph")
table(clusters)
sce.gse85241 <- computeSumFactors(sce.gse85241, min.mean=0.1, clusters=clusters)
summary(sizeFactors(sce.gse85241))
sce.gse85241 <- computeSpikeFactors(sce.gse85241, general.use=FALSE)
summary(sizeFactors(sce.gse85241, "ERCC"))
sce.gse85241 <- normalize(sce.gse85241)
```

### Identifying highly variable genes

We identify highly variable genes (HVGs) using `trendVar()` and `decomposeVar()`.
Again, we set `block=` to ensure that uninteresting differences between plates or donors do not inflate the variance.

```{r, fig.cap="Variance of normalized log-expression values for each gene in the GSE85241 dataset, plotted against the mean log-expression. The blue line represents the mean-dependent trend fitted to the variances of the spike-in transcripts (red). The small apparent discrepancy in the fit is caused by the fact that the trend is fitted robustly to the block-wise variances of the spike-ins, while the variances shown are averaged across blocks and not robust to outliers."}
block <- paste0(sce.gse85241$Plate, "_", sce.gse85241$Donor)
fit <- trendVar(sce.gse85241, block=block, parametric=TRUE) 
dec <- decomposeVar(sce.gse85241, fit)
plot(dec$mean, dec$total, xlab="Mean log-expression", 
	ylab="Variance of log-expression", pch=16)
is.spike <- isSpike(sce.gse85241)
points(dec$mean[is.spike], dec$total[is.spike], col="red", pch=16)
curve(fit$trend(x), col="dodgerblue", add=TRUE)
```

We order genes by decreasing biological component, revealing some usual suspects such as insulin and glucagon.

```{r}
dec <- cbind(Symbol=rowData(sce.gse85241)$Symbol, dec)
dec[head(order(dec$bio, decreasing=TRUE)),]
```

# References

